Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-08 main.tex
more of intro

\newcommand*{\LK}{\mathsf{LK}}
\newcommand*{\LKm}{\mathsf{LK + mix}}
\newcommand*{\MLL}{\mathsf{MLL}}
\newcommand*{\FOMLL}{\mathsf{MLL1}} 
\newcommand*{\FOMLLm}{\mathsf{MLL1 + mix}}

@@

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as ``syntax-free'' notion of proof, and as a possible solution
to Hilbert's 24th problem~\cite{hughes:invar} (see also~\cite{str:hilbert:24}).

>>>>

\newcommand*{\LK}{\mathsf{LK}}
\newcommand*{\LKm}{\mathsf{LK + mix}}
\newcommand*{\MLL}{\mathsf{MLL}}
\newcommand*{\FOMLL}{\mathsf{MLL1}} 
\newcommand*{\FOMLLm}{\mathsf{MLL1 + mix}}
\newcommand*{\FOKS}{\mathsf{KS1}}

@@

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as ``syntax-free'' notion of proof, and as a possible solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in the proof and consider the
proof as a combinatorial object, more precisely as a special kind
homomorphism between two graphs obeying certain properties.
\lutz{an example here?}

It has been shown by several authors how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This allows to
define a natural notion of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.

Recently, Acclavio and Stra\ss burger extended this notion to relevant
logics~\cite{acc:str:relevant} and to modal
logics~\cite{acc:str:modal}; and Heijlties, Hughes and Stra\ss burger
have provided combinatorial proofs for intuitionistic propositional
logic~\cite{HHS:lics19}.

In this paper we would like to push forward the idea that
combinatorial proofs can also for first order logic be used as a
notion of proof identity. \emph{First-order combinatorial proofs} have
been introduced by Hughes in~\cite{hughes:fopws}. But even though
Hughes shows that the conclusion of every first-order combinatorial
proof is a valid formula, his proof is not really satisfactory, as (i)
it is long and cumbersome, and (ii) it does not allow to read back a
syntactic proof based inference rules. In fact, there is the
fundamental problem that not all combinatorial proofs can be obtained
as translations of sequent calculus proofs.

In this paper we solve this issue by moving to a deep inference
system. More precisely, we introduce a new proof system, called
$\FOKS$, for first-order logic, that (i) reflects every combinatorial
proof, i.e., there is a surjective mapping from proofs in $\FOKS$ to
combinatorial proofs, that (ii) allows to provide simpler proofs of
soundness and completeness of combinatorial proofs, and (iii) admits
new decomposition theorems establishing a precise correspondence
between certain syntactic inference rules and certain combinatiorial
notions.

In general, a \emph{decomposition theorem} provides normal forms of
proofs, separating subsets of infernence rules of a proof system. A
prominent example of a decomposion theorem is Herbrand's
theorem~\cite{herbrand:phd}, which allows a separation between the
proposotional part and the quantifier part in a first-order
proof~\cite{gentzen:35:II,brunnler:06:herbrand}. Through the advent of
deep inference, new kinds of proof decompositions became possible,
most notably the separation between the linear part of a proof and the
resource management of a proof. It has been shown by
Stra{\ss}burger~\cite{str:07:RTA} that a proof in classical
propositional logic can be decomposed into a proof of (multiplicative)
linear logic, followed by a proof consisting only of contractions and
weakenings. In this paper we show that the same is possible for
first-order logic.

In summary, this paper contains the following contributions:
